### 2025-03-02 sec/0_abstract.tex
minor rewrites

\begin{abstract}
Accurate single cell detection in brightfield microscopy is crucial for biological research, yet data scarcity and annotation bottlenecks limit the progress of deep learning methods.
We investigate the use of unconditional models to generate synthetic brightfield microscopy images and evaluate their impact on object detection performance.
A U-Net based diffusion model was trained and used to create datasets with varying ratios of synthetic and real images.
Experiments with YOLOv8, YOLOv9 and RT-DETR reveal that training with synthetic data achieves comparable, and in some cases, slightly improved detection accuracies at lower Intersection over Union (IoU) thresholds.
A human expert survey demonstrates the high realism of generated images, with experts struggling to distinguish them from real microscopy images (accuracy $\sim$50\%).
Our findings suggest that diffusion-based synthetic data generation is a promising avenue for augmenting real datasets in microscopy image analysis, reducing the reliance on extensive manual annotation and potentially improving the robustness of cell detection models.
\end{abstract}

>>>>

\begin{abstract}
Accurate single cell detection in brightfield microscopy is crucial for biological research, yet data scarcity and annotation bottlenecks limit the progress of deep learning methods.
We investigate the use of unconditional models to generate synthetic brightfield microscopy images and evaluate their impact on object detection performance.
A U-Net based diffusion model was trained and used to create datasets with varying ratios of synthetic and real images.
Experiments with YOLOv8, YOLOv9 and RT-DETR reveal that training with synthetic data can achieve improved detection accuracies (at minimal costs).
A human expert survey demonstrates the high realism of generated images, with experts not capable to distinguish them from real microscopy images (accuracy $\sim$50\%).
Our findings suggest that diffusion-based synthetic data generation is a promising avenue for augmenting real datasets in microscopy image analysis, reducing the reliance on extensive manual annotation and potentially improving the robustness of cell detection models.
\end{abstract}

### 2025-03-02 sec/2_relatedwork.tex
minor rewrites

The advent of deep learning, particularly Generative Adversarial Networks (GANs), marked a significant shift in synthetic image generation.
GANs offered the potential to learn complex data distributions directly from real images, leading to more realistic and diverse synthetic ouputs.
In the context of microscopy, GANs have been extensively explored for various tasks.
Cross-modality translation, aiming to predict fluorescence microscopy images from brightfield inputs, has been a prominent application.
Christiansen \etal introduced ``In Silico Labeling'' (ISL) using deep learning to predict fluorescence labels from transmitted light images~\cite{christiansen_silico_2018}.
Building upon this, Lee \etal developed DeepHCS and DeepHCS++ to transform brightfield images into multiple fluorescence channels relevant for high-contrast screening, leveraging adversarial losses for realism~\cite{gyuhyun_lee_deephcs_2018,gyuhyun_lee_deephcs_2021}.
GANs have also been applied to enhance image quality achieving super-resolution under a large field-of-view~\cite{zhang_high-throughput_2019} and generate isolate cell images approximating the shape and texture distributions of cell components~\cite{marin_scalbert_generic_2019}.
Despite these success, GANs are known to suffer from training instability and mode collapse, limiting the diversity and reliability of generated samples, particularly when training data is scarce, complex or imbalanced, as often encountered in microscopy~\cite{juan_c_caicedo_evaluation_2019,ricard_durall_combating_2020}.

@@

While diffusion models are increasingly used in microscopy image generation, research on their direct application for enhancing \textbf{object detection}, specifically in the challenging domain of brightfield microscopy, remains significantly limited.
Existing studies primarily focus on image generation quality evaluation and downstream tasks like segmentation or image-based profiling.
The critical question of whether and how diffusion-based synthetic brightfield microscopy images can effectively improve the performance of object detection models for single cell detection has not been systematically addressed.
This work directly addresses this gap by undertaking a comprehensive evaluation of the impact on diffusion-based synthetic brightfield microscopy image data on single cell detection performance.
By focusing on object detection, this research contributes a novel perspective to the field, exploring the practical utility of synthetic data for a fundamental task in quantitative microscopy.

>>>>

The advent of deep learning, particularly Generative Adversarial Networks (GANs), marked a significant shift in synthetic image generation.
GANs offered the potential to learn complex data distributions directly from real images, leading to more realistic and diverse synthetic ouputs.
In the context of microscopy, GANs have been extensively explored for various tasks.
Cross-modality translation, aiming to predict fluorescence microscopy images from brightfield inputs, has been a prominent application.
Christiansen \etal introduced ``In Silico Labeling'' (ISL) using deep learning to predict fluorescence labels from transmitted light images~\cite{christiansen_silico_2018}.
Building upon this, Lee \etal developed DeepHCS and DeepHCS++ to transform brightfield images into multiple fluorescence channels relevant for high-contrast screening, leveraging adversarial losses for realism~\cite{gyuhyun_lee_deephcs_2018,gyuhyun_lee_deephcs_2021}.
GANs have also been applied to enhance image quality achieving super-resolution under a large field-of-view~\cite{zhang_high-throughput_2019} and generate isolate cell images approximating the shape and texture distributions of cell components~\cite{marin_scalbert_generic_2019}.
Despite these successes, GANs are known to suffer from training instability and mode collapse, limiting the diversity and reliability of generated samples, particularly when training data is scarce, complex or imbalanced, as often encountered in microscopy~\cite{juan_c_caicedo_evaluation_2019,ricard_durall_combating_2020}.

@@

While diffusion models are increasingly used in microscopy image generation, research on their direct application for enhancing \textbf{object detection}, specifically in the challenging domain of brightfield microscopy, remains significantly limited.
Existing studies primarily focus on image generation quality evaluation and downstream tasks like segmentation or image-based profiling.
The critical question of whether and how diffusion-based synthetic brightfield microscopy images can effectively improve the performance of object detection models for single cell detection has not been systematically addressed.
This work is directly pertinent to the aforementioned issue by undertaking a comprehensive evaluation of the impact on diffusion-based synthetic brightfield microscopy image data on single cell detection performance.
By focusing on object detection, this research contributes a novel perspective to the field, exploring the practical utility of synthetic data for a fundamental task in quantitative microscopy.

### 2025-03-02 sec/3_methodology.tex
minor rewrites

The diffusion model was trained using a dataset of 10,000 patches ($512 \times 512$) extracted from real brightfield microscopy images.
These images were acquired using a CELLAVISTA 3.1 RS HE high-throughput microscope (Synentec GmbH) with a $10\times$ objective lense, resuting in 8-bit grayscale images.
The source material consisted of stably transfected CHO cells (CHO-K1 and CHO DG44 variants) expressing eGFP, imaged in 96-well plates at three different densities (300, 150, and 1 cell per well).

@@

We employed an unconditional diffusion models with a U-Net architectures and attention mechanisms as the backbone for image generation.
The U-Net architecture was chosen for its proven effectiveness in image generation and segmentation tasks, particularly in biomedical imaging, due to its ability to capture both local and global context through its encode-decoder structure and skip connections~\cite{ronneberger_u-net_2015}.
The specific U-Net architecture implemented consisted of downsampling and upsampling blocks with residual connections to facilitate gradient flow and improve training stability.
Attention mechanisms were optionally incorporated into both the downsampling and upsampling paths in some model variants to investigate their impact on image generation quality.
The number of up and down blocks, channel configurations and the placement of attention blocks were varied across different model configurations to evaluate architectural choices.

>>>>

The diffusion model was trained using a dataset of 10,000 patches ($512 \times 512$) extracted from real brightfield microscopy images.
These images were acquired using a CELLAVISTA 3.1 RS HE high-throughput microscope (Synentec GmbH) with a $10\times$ objective lense, resulting in 8-bit grayscale images.
The source material consisted of stably transfected CHO cells (CHO-K1 and CHO DG44 variants) expressing eGFP, imaged in 96-well plates at three different densities (300, 150, and 1 cell per well).

@@

We employed an unconditional diffusion models based on U-Net architectures and attention mechanisms as the backbone for image generation.
The U-Net architecture was chosen for its proven effectiveness in image generation and segmentation tasks, particularly in biomedical imaging, due to its ability to capture both local and global context through its encode-decoder structure and skip connections~\cite{ronneberger_u-net_2015}.
The specific U-Net architecture implemented consisted of downsampling and upsampling blocks with residual connections to facilitate gradient flow and improve training stability.
Attention mechanisms were optionally incorporated into both the downsampling and upsampling paths in some model variants to investigate their impact on image generation quality.
The number of up and down blocks, channel configurations and the placement of attention blocks were varied across different model configurations to evaluate architectural choices.

### 2025-03-02 sec/4_experimentsandresults.tex
minor rewrites

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{images/norm_confusion_matrix}
    \caption{Normalized confusion matrix of expert survey classifications. Overall accuracy is approximately 50\%, demonstrating the significant difficulty for experts in distinguishing real and synthetic brightfield microscopy images.}
    \label{fig:conf_matrix}
\end{figure}

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, slightly improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, marginally exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, very slightly outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
These results suggest that synthetic data augmentation can maintain or even slightly enhance detection performance at lower IoU thresholds, potentially due to increased data variability and improved model robustness.

However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease merges for models trained with increasing proportions of synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.

Qualitative assessment of sample detections (examples in supplementary material) revealed that all models, including those trained with synthetic data, generally performed well in detecting cells under diverse conditions, including overlapping cells and cells at image boundaries.
However, subtle differences in detection precision and confidence scores were observed, warranting further investigation into the fine-grained impact of synthetic data on model behavior.

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=0.75\linewidth]{images/norm_confusion_matrix}
    \caption{Normalized confusion matrix of expert survey classifications. Overall accuracy is approximately 50\%, demonstrating the significant difficulty for experts in distinguishing real and synthetic brightfield microscopy images.}
    \label{fig:conf_matrix}
\end{figure}

\subsection{Object Detection Performance}
\label{subsec:object-detection-performance}
Table~\ref{tab:model-performances} summarizes the object detection performance, presenting mAP metrics for all trained models across the seven datasets.
Analyzing mAP\@50, we observe that models trained with synthetic data replacing parts of the real data (\textit{scc\_10}, \textit{scc\_30}, \textit{scc\_50}) achieve comparable, and in some cases, improved performance compared to models trained solely on real data (\textit{scc\_real}).
Specifically, the YOLOv8s model trained on the \textit{scc\_30} dataset achieved a mAP\@50 of 0.9043, marginally exceeding the 0.8947 achieved by the same model trained on the \textit{scc\_real} dataset.
Similarly, the RT-DETR-l model trained on the \textit{scc\_10} dataset reached a mAP\@50 of 0.9147, outperforming the 0.9146 achieved on the \textit{scc\_real} dataset.
These results suggest that synthetic data augmentation can maintain or even enhance detection performance at lower IoU thresholds, potentially due to increased data variability and improved model robustness.

However, when examining performance at higher IoU thresholds, specifically mAP\@75 and mAP\@50:95, a subtle trend of performance decrease emerges for models trained with increasing proportions of synthetic data.
This trend is more pronounced for the RT-DETR models, indicating a potentially greater sensitivity of transformer-based architectures to the characteristics of synthetic data.

Qualitative assessment of sample detections revealed that all models, including those trained with synthetic data, generally performed well in detecting cells under the conditions proposed by the test split, including overlapping cells and cells at image boundaries.
However, subtle differences in detection precision and confidence scores were observed, warranting further investigation into the fine-grained impact of synthetic data on model behavior.

### 2025-03-02 sec/5_discussion.tex
minor rewrites

Object detection experiments further reveal the practical utility of diffusion-based synthetic data for single cell detection.
The comparable, and in some cases, slightly improved performance of models trained with synthetic data augmentation at mAP\@50 demonstrates that synthetic images can effectively capture essential cell features and spatial distributions relevant for accurate cell localization.
This directly addresses our main research question and suggests that synthetic data can be a valuable asset in training robust cell detection models, particularly when labeled real data is limited.
The slight performance enhancement observed at mAP\@50 for some models might be attributed to the increased variability introduced by the synthetic data generation process, potentially improving model generalization and robustness to variations in real-world microscopy images.

>>>>

Object detection experiments further reveal the practical utility of diffusion-based synthetic data for single cell detection.
The comparable, and in some cases, improved performance of models trained with synthetic data augmentation at mAP\@50 demonstrates that synthetic images can effectively capture essential cell features and spatial distributions relevant for accurate cell localization.
This directly addresses our main research question and suggests that synthetic data can be a valuable asset in training robust cell detection models, particularly when labeled real data is limited.
The slight performance enhancement observed at mAP\@50 for some models might be attributed to the increased variability introduced by the synthetic data generation process, potentially improving model generalization and robustness to variations in real-world microscopy images.

### 2025-03-02 sec/6_conclusion.tex
minor rewrites

\section{Conclusion}
\label{sec:conclusion}
This paper provides a comprehensive investigation into the use of diffusion-based synthetic brightfield microscopy images for enhancing single cell detection.
Our expert survey demonstrates the remarkable realism of diffusion-generated images, achieving near-indistinguishability from real microscopy acquisitions.
Object detection experiments reveal that models trained with synthetic data achieve comparable, and in some cases, slightly improved performance to real-data training, particularly for simpler cell localization (mAP\@50).
While subtle limitations exist in  replicating fine cell boundary details and achieving optimal performance at higher IoU thresholds, our findings strongly highlight the promise of diffusion-based synthetic data generation as a valuable tool for microscopy image analysis.
This approach offers a pathway to address data scarcity, reduce annotation burdens, and potentially improve the robustness and accessibility of advanced cell detection techniques in biological and medical research.
Future research directions should focus on refining diffusion models for brightfield microscopy image generation, improving the fidelity in capturing fine cellular details, exploring conditional generation strategies for broader applicability across diverse microscopy modalities and biological contexts, and further investigating the optimal strategies for integrating synthetic data into training pipelines to maximize the benefits for cell detection and other microscopy image analysis tasks.

>>>>

\section{Conclusion}
\label{sec:conclusion}
This paper provides a comprehensive investigation into the use of diffusion-based synthetic brightfield microscopy images for enhancing single cell detection.
Our expert survey demonstrates the remarkable realism of diffusion-generated images, achieving near-indistinguishability from real microscopy acquisitions.
Object detection experiments reveal that models trained with synthetic data achieve comparable, and in some cases, improved performance to real-data training, particularly for simpler cell localization (mAP\@50).
While subtle limitations exist in  replicating fine cell boundary details and achieving optimal performance at higher IoU thresholds, our findings strongly highlight the promise of diffusion-based synthetic data generation as a valuable tool for microscopy image analysis.
This approach offers a pathway to address data scarcity, reduce annotation burdens, and potentially improve the robustness and accessibility of advanced cell detection techniques in biological and medical research.
Future research directions should focus on refining diffusion models for brightfield microscopy image generation, improving the fidelity in capturing fine cellular details, exploring conditional generation strategies for broader applicability across diverse microscopy modalities and biological contexts, and further investigating the optimal strategies for integrating synthetic data into training pipelines to maximize the benefits for cell detection and other microscopy image analysis tasks.
